---
title: "Chess ratings"
author: "Luis Munoz Grass"
date: "2024-09-21"
output: html_document
---

For this project we’re given a text file with chess tournament results where the information has some structure. The idea is to create an R Markdown file that generates a .CSV file with the following information for all of the players:
Player’s Name
Player’s State
Total Number of Points
Player’s Pre-Rating
Average Pre Chess Rating of Opponents

For the first player, the information would be:
Gary Hua, ON, 6.0, 1794, 1605

1605 was calculated by using the pre-tournament opponents’ ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and dividing by the total number of games played.

```{r, eval=TRUE, message=FALSE, warning=FALSE}

library(stringr)
```

## Import data from tournament info text file.

```{r, eval=TRUE, message=FALSE, warning=FALSE}
chesstour <- read.csv(paste0("https://raw.githubusercontent.com/Lfirenzeg/msds607labs/refs/heads/main/tournamentinfo.txt"), header=FALSE)
```

```{r, eval=TRUE, message=FALSE, warning=FALSE}
## Remove header rows.
chesstour <- chesstour[-c(1:4),]
```


We can use head() to check if the first 4 rows were removed successfully.
```{r, eval=TRUE, message=FALSE, warning=FALSE}
head(chesstour)
```


```{r, eval=TRUE, message=FALSE, warning=FALSE}
## Select sequential pieces for each player info and its rating into different data frames.
PlyInfo <- chesstour[seq(1,length(chesstour),3)]
RtiInfo <- chesstour[seq(2,length(chesstour),3)]
```


```{r, eval=TRUE, message=FALSE, warning=FALSE}
## Further extract required elements for this project.
  #### Extract player's number and convert its character type to integer.
  Player_Number <- unlist(str_extract(PlyInfo,"[[:digit:]]{1,2}"))
  Player_Number <- as.integer(Player_Number)
```

  
```{r, eval=TRUE, message=FALSE, warning=FALSE}
  #### Extract player's name.
  Player_Name <- unlist(str_extract(PlyInfo,"[[:alpha:]]{2,}.{1,}([[:alpha:]]{2,})"))
```


```{r, eval=TRUE, message=FALSE, warning=FALSE}    
  #### Extract player's points and convert its character type to numeric.
  Player_Points <- unlist(str_extract(PlyInfo,"[[:digit:]]{1}\\.[[:digit:]]{1}"))
  Player_Points <- as.numeric(Player_Points)
```
  
  
```{r, eval=TRUE, message=FALSE, warning=FALSE}
  #### Extract player's state.
  Player_State <- unlist(str_extract(RtiInfo,"[[:alpha:]]{1,2}"))
```  



```{r, eval=TRUE, message=FALSE, warning=FALSE}  
  #### Extract player's pre-rating and using sub function to find and remove ": " and then convert to        numeric.
  Player_Rating <- unlist(str_extract(RtiInfo, ":\\s{1,}[[:digit:]]{3,4}"))
  Player_Rating <- sub(":\\s{1,}","",Player_Rating)
  Player_Rating <- as.numeric(Player_Rating)
```
  
  
```{r, eval=TRUE, message=FALSE, warning=FALSE}    
  #### Extract each player's opponents' number.
  Opponents_Number <- str_extract_all(PlyInfo, "[[:digit:]]{1,}\\|")
  Opponents_Number <- str_extract_all(Opponents_Number, "[[:digit:]]{1,}")
```
  
  
```{r, eval=TRUE, message=FALSE, warning=FALSE}    
  #### Calculate the sum of Win, Lose, and Draw counts for each player.
  Player_GamesCount <- str_count(PlyInfo, "\\|W") + str_count(PlyInfo, "\\|L") + str_count(PlyInfo, "\\|D")
```
  
      
  
```{r, eval=TRUE, message=FALSE, warning=FALSE}
  #### Calculate the sum of opponents' rating for each player.
  Player_TotalRating <- length(Player_Number)
  
   for (i in 1:length(Player_Number))
   { 
          Player_TotalRating[i] <-  sum(Player_Rating[as.numeric(unlist(Opponents_Number[Player_Number[i]]))])
          
   }
```


```{r, eval=TRUE, message=FALSE, warning=FALSE}     
  #### Calculate the average pre chess rating of opponents for each player.
  Player_AverageRating <- round(Player_TotalRating/Player_GamesCount,0)
```

  
```{r, eval=TRUE, message=FALSE, warning=FALSE} 
## Combine above data frames.
chesstour_summary <- cbind.data.frame(Player_Name,Player_State,Player_Points,Player_Rating, Player_AverageRating)
```


```{r, eval=TRUE, message=FALSE, warning=FALSE} 
## Update the columns' names.
colnames(chesstour_summary) <- c("Player’s Name", "Player’s State", "Total Number of Points", "Player’s Pre-Rating", "Average Pre Chess Rating of Opponents")
```


```{r, eval=TRUE, message=FALSE, warning=FALSE}
## Display few rows of the summary results.
head(chesstour_summary)
```


```{r, eval=TRUE, message=FALSE, warning=FALSE}   
## Export the summary to a CSV file.
write.csv(chesstour_summary,'ChesstourSummary.csv')
```