---
title: "Project 3 - Job Posts skills"
author: "   "
date: "2024-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tyding Skills from Job Posts

Once we have a .csv file containing the list of skills for each job post we can proceed to organize the information pulled with Pythong a bit more.

Load the needed libraries

```{r load-packages, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library (stringr)
```

And load the data, which in this case is hosted in GitHub

```{r, get data from GitHub}
skills_url <- "https://raw.githubusercontent.com/Lfirenzeg/msds607labs/refs/heads/main/Project3/skills.csv"  
companies_skills_url <- "https://raw.githubusercontent.com/Lfirenzeg/msds607labs/refs/heads/main/Project3/companies_skills.csv"

# Load the CSV files into R data frames
skills_df <- read.csv(skills_url, stringsAsFactors = FALSE)
companies_skills_df <- read.csv(companies_skills_url, header = FALSE, stringsAsFactors = FALSE)

# Display the first few rows of each table to confirm data was loaded succesfully 
head(skills_df)
head(companies_skills_df)
```

Next, we'll assign headers to the companies_skills_df table and remove some columns that are not necessary for the analysis, such as URL and Timestamp

```{r, assign headers and remove 2 columns}
# To assign column names to the companies_skills_df table
colnames(companies_skills_df) <- c("URL", "Timestamp", "Skills", "Position", "Company", "Industry")

# And then remove the "URL" and "Timestamp" columns using dplyr's select function
companies_skills_df <- companies_skills_df %>%
  select(-URL, -Timestamp)

```

To organize the data a bit more, we'll be separating the companies_skills_df table into two separate tables—one focused on company information and the other on skills
```{r, splitting a table}
# Reordering the columns using dplyr's select function
companies_skills_df <- companies_skills_df %>%
  select(Position, Skills, Company, Industry)

# Table 1: Skills Job Post (without Industry)
skills_job_post <- companies_skills_df %>%
  select(Position, Skills, Company)

# Table 2: Company and Industry Table
company_industry_table <- companies_skills_df %>%
  select(Company, Industry) %>%
  distinct()  # using distinct to ensure there's no duplicated company-industry rows

# Saving the new tables to CSV files
write.csv(skills_job_post, "skills_table.csv", row.names = FALSE)
write.csv(company_industry_table, "company_industry_table.csv", row.names = FALSE)
```

Now let's add the additional information to skills_job_post, such as how many skills are listed in total for each job post.

```{r, add column total skill count}
skills_job_post <- skills_job_post %>% 
  mutate(Total_Skills = str_count(Skills, ",") + 1)  # Counting commas and adding 1 to get the total count
```

And also include additional columns for each type of skill. To do this the idea is to use the skills_df table that has all the original skills classified by type. A function will be used for this extracting the information from skills_df.

```{r, total skills by type}
# Create a new column to replace the skills with categories
skills_job_post <- skills_job_post %>%
  rowwise() %>%
  mutate(skills_replaced_by_category = str_replace_all(Skills, setNames(skills_df$category, skills_df$skill)))

# Now split the 'skills_replaced_by_category' by commas to get individual categories
skills_job_post <- skills_job_post %>%
  mutate(
    Programming_Languages = str_count(skills_replaced_by_category, "Programming Languages"),
    Technical_Skills = str_count(skills_replaced_by_category, "Technical Skills"),
    Business_Skills = str_count(skills_replaced_by_category, "Business Skills"),
    Soft_Skills = str_count(skills_replaced_by_category, "Soft Skills")
  )

# Remove the temporary column (optional)
skills_job_post <- skills_job_post %>%
  select(-skills_replaced_by_category)

# View the updated dataframe
head(skills_job_post)
```

